
% vim: set tw=79 sw=3 expandtab:

\documentclass[nobib]{tufte-book}

\input{preamble}

\begin{document}

\input{front-matter}

%%
% Start the main matter (normal chapters)
\mainmatter

\chapter{Distance}
\label{ch:distance}

\newthought{A point}, as introduced in high-school geometry, is a particular,
pure location, without any extent: no length, no width, no height.  An object
such as a curve, a surface, or a volume has points in it.

There are, however, problems that can arise from imagining, say, a curve as
\emph{merely} a collection of points.  A point has no length.  Two points could
touch only if they were in the same location, but then they would be the same
point.  So no two distinct points can be in contact.  There is no arrangement
of points, lying in unbroken contact, such that they constitute a length.
Also, in introductory physics, the student learns to involve physical dimension
in calculation.  Two quantities cannot be added together if they differ in
dimension, and if two quantities agree in dimension, their sum must be of the
same dimension.\sidenote[][-1in]{%
   For example, one cannot evaluate the sum of a distance and a mere number, as
   in $3 \: \text{m}^2 + 2$, where ``$\text{m}$'' indicates ``meter''.  The
   first term has the dimension of area, and the second term is dimensionless.
   Also, the result of $2 + 5$ cannot be $7 \: \text{m}$.}
A point has no extension in the dimension of length, and so even if two
distinct points could touch, there is no way to add up any property in each
point to result in a length.

\begin{marginfigure}
   \begin{center}
   \begin{asy}
      defaultpen(fontsize(8pt));
      settings.prc = false;
      size(4cm,0);
      import three;
      real step = 3*pi/16;
      triple p0 = cos(0*step)*X + sin(0*step)*Y;
      triple p1 = cos(1*step)*X + sin(1*step)*Y;
      triple p2 = cos(2*step)*X + sin(2*step)*Y;
      triple p3 = cos(3*step)*X + sin(3*step)*Y;
      triple p4 = cos(4*step)*X + sin(4*step)*Y;
      path3 a = arc(c=O, v1=p0, v2=p4);
      currentprojection=orthographic(camera=Z);
      draw(a);
      draw(p0--p2--p4, blue);
      draw(p0--p1--p2--p3--p4, red);
   \end{asy}
   \end{center}
   \caption{%
      Convergence toward limiting length of arc.  The integral of the length
      along the curved path is the limit of a sequence of approximate sums.
      The sum of the red distances is a better approximation than the sum of
      the blue distances. The smooth arc is not composed of an infinite number
      of straight segments. That would be a contradiction; a curve is not
      straight.  Rather, the \emph{length} of the curved arc is the limit of a
      sequence of sums of distances.  No sum will in general be equal to the
      length along the curve, but the limit of the sequence of sums can be
      found.%
   }
   \label{fig:arc}
\end{marginfigure}

\newthought{A solution} is found by combining calculus with the idea of the
\emph{distance} between two points.  The length along a curve is broken down
not into points but into ever smaller distances.  In physical space, there is a
unique distance between any two points.  The length along a curve is defined as
the limit of a sequence of sums.  See Figure~\ref{fig:arc}.  Each subsequent
sum in the sequence is the sum a larger number of smaller distances.  The
dimension of the addends is the dimension of the sum.  So long as the curve be
\emph{smooth}, the limit of the sequence of the sums will be the length of the
curve.  A curve is smooth if, for any two points sufficiently close together
along the curve, the difference between their separation distance and the
length between them along the curve is as small as desired.

Just as a length is properly composed of distances, a surface must be composed
of areas, and a space must be composed of volumes.  In turn, the idea of area
and the idea of volume are built upon the idea of distance.

Regardless of the number of dimensions of length a space might have, it can
support basic notions of geometry only if there be a unique \emph{distance}
between every pair of points in the space.  A path through physical space is
not merely a collection of points but also a rule for associating a unique
distance between every pair of points on the path.

\section{Distance and Path}

\section{Geodesic}

\appendix

%%
% The back matter contains appendices, bibliographies, indices, glossaries,
% etc.
\backmatter

\input{fdl-1.3}

\bibliography{my}
\bibliographystyle{plainnat}

\printindex

\end{document}

