
% vim: set tw=79 sw=3 expandtab:

\chapter{Distance}
\label{ch:distance}

A point, as introduced in high-school geometry, is a pure location, without any
\emph{extent}.  Each of
\begin{itemize}[noitemsep]
   \item a path (a one-dimensional space),
   \item a surface (a two-dimensional space),
   \item a proper space (a three-dimensional space), and
   \item a hyperspace (a space with four or more dimensions)
\end{itemize}
has points in it.  Within each kind of space, one can conceive of an
\emph{extended} object, an object that has extent along at least one dimension.
For example, a segment in a one-dimensional space, a triangle in a
two-dimensional space, or a cube in a three-dimensional space.

An extent is an unbroken \emph{length} corresponding to a thing's size.  What
we call ``length'' is the principle of extension.  A geometric object might be
extended in multiple ways, as in width and in height.  Each of the width and
the height is fundamentally a length, though each is a length in a different
dimension.  For example, consider a $120^\circ$~circular arc in the Euclidean
plane.  There is a length along the arc, which is a segment in the
one-dimensional space of the circle on which the arc sits.  In the space of the
plane, the arc has extension in two dimensions.
\begin{figure}
   \begin{center}
   \begin{asy}
      defaultpen(fontsize(8pt));
      settings.prc = false;
      size(9.5cm,0);
      import three;
      real step  = pi/12;
      real phase = pi/6;
      real yoff  = sin(0*step + phase);
      triple p00 = cos(0*step + phase)*X + (sin(0*step + phase) - yoff)*Y;
      triple p01 = cos(1*step + phase)*X + (sin(1*step + phase) - yoff)*Y;
      triple p02 = cos(2*step + phase)*X + (sin(2*step + phase) - yoff)*Y;
      triple p03 = cos(3*step + phase)*X + (sin(3*step + phase) - yoff)*Y;
      triple p04 = cos(4*step + phase)*X + (sin(4*step + phase) - yoff)*Y;
      triple p05 = cos(5*step + phase)*X + (sin(5*step + phase) - yoff)*Y;
      triple p06 = cos(6*step + phase)*X + (sin(6*step + phase) - yoff)*Y;
      triple p07 = cos(7*step + phase)*X + (sin(7*step + phase) - yoff)*Y;
      triple p08 = cos(8*step + phase)*X + (sin(8*step + phase) - yoff)*Y;
      currentprojection=orthographic(camera=Z);
      // ".." indicates bezier curve.
      path3 a = p00..p01..p02..p03..p04..p05..p06..p07..p08;
      draw(a, gray);
      pen b8818 = blue  + linetype(new real[]{8,8,1,8});
      pen g4414 = green + linetype(new real[]{4,4,1,4});
      pen r2212 = red   + linetype(new real[]{2,2,1,2});
      draw(p00--p04--p08, b8818);
      draw(p00--p02--p04--p06--p08, g4414);
      draw(p00--p01--p02--p03--p04--p05--p06--p07--p08, r2212);
   \end{asy}
   \end{center}
   \caption{A $120^\circ$~arc and approximations to its length.}
   \label{fig:arc}
\end{figure}
See Figure~\ref{fig:arc}.

Any extended object will have within it some of the points that are in the
space containing the object.  Yet because a point has no extent, an extended
object does not have its extension by virtue of the points themselves.  Rather,
extension is defined as a relationship between points.

\section{Shortest Path}

Between any two points in a space, there is a special length called a
\emph{distance}; contained in this idea is, as it turns out, most of what we
mean by ``space.'' The distance between any two points in a space is the length
of the shortest path between them.

For a one-dimensional space, the existence of a shortest length between any
pair of points implies that the space itself is a single, unbroken extent.

For a space of two or more dimensions, however, the assertion of a distance
between any pair of points implies that there are infinitely many unbroken
paths between the two points and that one of those paths has the shortest
length.

\section{Length Along Curved Path}

The length along a curved path is the \emph{limit} of a sequence of sums.

In order to find the first sum in the sequence, suppose that, in addition to
the point at the beginning of the path and the point at the end of the path, we
pick a collection of points along the path so that the distance between one
point and the next in the collection is the same for every point.  Starting at
one end of the path, we can add up the distance from one point to the next
until we arrive at the other end of the path.  Because the distance between one
point and the next corresponds to the shortest length between them, the sum of
the distances will be less than the length along the path if the path be
curved.  See, for example, the blue, dot-dashed segments in
Figure~\ref{fig:arc}.

We find the second sum by picking, for each pair of consecutive points in the
first collection, a new point along the path between them and equidistant from
them.  Thus augmenting our collection of points, we again start at one end of
the path and add up the distance from one point to the next until we arrive at
the other end of the path.  See the green segments in Figure~\ref{fig:arc}.

If the path be \emph{smooth}\footnote{%
   A path is smooth if, for any two points sufficiently close together along
   the path, the fractional difference $\frac{\lambda - d}{\lambda}$ between
   \begin{itemize}[noitemsep]
      \item the length $\lambda$ along the path between the points and
      \item their separation distance $d$
   \end{itemize}
   can be made so close to zero as desired.%
}
and not straight, then, continuing to augment our collection of points as we
have done above, we find that the sum at every stage increases, but the
sequence of sums approaches a fixed upper limit $L$.  Regardless of how many
sums we generate, the last sum is always less than the limit, but, for any
length $\ell$ smaller than $L$---even if $L - \ell$ be so tiny as we might
imagine---we can always keep augmenting our collection until the sum is greater
than $\ell$.  Therefore, the limit $L$ is the least upper bound of the sequence
of sums and in fact the length along the path.\footnote{%
   The existence of the limit does not require that the curved path be regarded
   as composed of a large number of straight segments. In fact, the limit is a
   length greater than that of any approximating collection of straight paths.
   Yet the limit exists and is computable for any shape of smooth curve.%
}

\section{Principle of Extension}

Any length is either a sum of distances or the limit of a sequence of such
sums. A surface area, too, or a volume ultimately depends on distance because
it is a product of lengths.  The idea of area and the idea of volume are built
upon the idea of length and therefore, ultimately, upon the idea of distance.

Regardless of the number of dimensions a space might have, it can support the
basic notion of extended geometry only if there be a unique \emph{distance}
between every pair of points in the space.  A path of definite length is not
merely a collection of points but also a rule for associating a unique distance
between every pair of points on the path.  The idea of distance is the
principle of continuous, unbroken geometry.  A continuous length is properly
divided not into points but rather into infinitesimal, contiguous elements of
distance.\footnote{%
   There is a problem that can arise from imagining a path (for example) as
   \emph{merely} a collection of points.  Because a point has no length, no
   point can touch another point.  Regardless of how close together two points
   might be, each has zero length, and so there is a gap between them.  Because
   no point can touch another point, there is no arrangement of points, lying
   in unbroken contact, such that they constitute a length.  The idea of an
   unbroken length, a continuously extended path, is more fundamental than the
   idea of a point:  A point is a pure location \emph{in} or \emph{on}
   something, such as a path, a surface, or proper space.  That thing must
   exist first, in order for there then to be points identified on it, and it
   must define what the distance between a pair of points is.  Beginning in the
   20th Century, a geometric object has often been considered as essentially a
   collection of points.  This can be reasonable only if one remember that
   there is also, somehow associated with the definition of the points, the
   idea of a \emph{distance} between every pair of points.%
}
This idea is what enables integral calculus to work.

\section{Geodesic}

